---
title: 'Previous Electoral Indicators: House Elections Data'
author: Luke Tomes
date: '2022-09-15'
slug: house-elections-data
categories: []
tags: []
authors: []
description: ''
externalLink: ''
series: []
---
```{r setup, include = FALSE, message = FALSE, echo = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(echo = TRUE, fig.align="center")
```

```{r, echo = FALSE, message = FALSE}
# Load Libraries
library(tidyverse)
library(usmap)
library(sf)
library(rmapshaper)
```

*This blog is part of a series of articles meant to progressively understand how election data can be used to predict future outcomes. I will add to this site on a weekly basis under the direction of [Professor Ryan D. Enos](<https://www.ryandenos.com>). In the run-up to the midterm elections on November 8, 2022, I will draw on all I have learned regarding what best forecasts election results, and I will predict the outcome of the NE-02 U.S. Congressional Election.*

## House of Representatives Election Data Analysis

In this inaugural blog, I am examining data on elections for the U.S. House of Representatives in the 2010's in order to better see how it can translate into an effective model for the 2022 midterm elections. A prominent question is how we can look at vote share margins and seat shares in congressional districts and use that data as a baseline to predict future election outcomes. By witnessing trends over time in a district's vote share margins and the capability of a district to elect an official of a certain party, one can plausibly predict the direction where that district is headed in the future.

### Party Vote Share Margins By State

One of the most fundamental aspects to the House of Representatives is its roots in being the federal legislative body closest to the citizenry. Every two years, 435 districts across the country elect a representative to carry their interests to Washington DC. Scrutinizing voting trends in these elections at district-wide, statewide, and nationwide levels is of great interest to those tapped into politics, because biennially there is new data on the country's ever-changing political temperature. In the visualization below, each state displays their respective vote share margin that the party who wins a majority of votes possesses. For instance, Wyoming boasts a hearty red color, indicating that Republicans win a much larger share of the total votes cast than Iowa with its faded red. Wyoming and California are known to be strongholds for Republicans and Democrats, respectively, but no matter how large a party's margin in any state, there are only a certain number of members they can elect to the House. Also, looking at the vote share margins at a district level is more informative, because districts can often be drawn to heavily favor one party, potentially contributing to a false notion that a party has full control over one state.

```{r, echo = FALSE, message = FALSE}
# load in data
popvote1 <- read_csv("house party vote share by district 1948-2020.csv")
```

```{r, echo = FALSE, message = FALSE, include = FALSE}
# need to filter out this year that contains 0 votes 
# will mess up new variables calculating margins
popvote1 %>%
  filter(raceYear == 2014,
         district_id == "OK01")
```

```{r, echo = FALSE,message = FALSE}
# Select for variables and create new one
# Exclude OK-01 Because Oklahoma) for 1b continuity?
popvote_margin <- popvote1 %>%
  filter(raceYear == 2014,
         district_id != 'OK01') %>%
  select(raceYear, State, DemVotes, RepVotes, district_num, district_id, WinnerParty) %>%
  group_by(State) %>%
  mutate(RepVoteMargin = (sum(RepVotes))/sum(RepVotes + DemVotes),
         DemVoteMargin = (sum(DemVotes))/sum(RepVotes + DemVotes),
         VoteMargin = RepVoteMargin - DemVoteMargin) %>%
  rename(state = State)
```

```{r, echo = FALSE, message = FALSE}
# plot US Map Vote Share Margin
state_map <- plot_usmap(data = popvote_margin,
                        regions = "states",
                        values = "VoteMargin",
                        labels = TRUE,
                        label_color = "black") +
  scale_fill_gradient2(low = "blue",
                       mid = "white",
                       high = "red",
                       name = 'Vote Share Margin') +
  labs(title = "U.S. Congressional Vote Margin by State, 2014",
       caption = "Data Provided by Professor Ryan Enos of Harvard University.",
       subtitle = "A positive margin indicates a GOP victory.\nA negative margin indicates a Democratic Party victory.\nA margin larger in magnitude indicates a more decisive victory.\n") +
  theme_void()

# Set labels font size
state_map$layers[[2]]$aes_params$size <- 2
print(state_map)
```

### Vote Margin by U.S. Congressional District 

Here we can further take a peek into how Democrats and Republicans perform on a district level. Oftentimes districts lack competition between the parties because of gerrymandering that works to assist a political party by lessening their probability of being defeated in the general election by the candidate of the opposing party. Gerrymandering often leads to heated primaries between members of a party since there is not as large of an incentive to win over independents or members of the opposing party, of which there are so few. From the previous visaulization, it would be easy to deduce that states with a large GOP vote share margin like Alabama would not elect a Democrat as a representative for their delegation to the House. However, upon examining Alabama on a district-level, we can see they did elect a Democrat in 2014 with strong enthusiasm. Gerrymandering is one way you can explain this: when the new district maps were being drawn, the GOP-sided officials in charge acknowledged there were a fair share of Democrats residing in the west side of Alabama. Instead of dispersing them across many districts, and allowing Republicans to become potentially more vulnerable in them, the GOP conceded only one district to the Democrats, thus saving the many others for themselves. Examining how these districts change in shape and in how voters vote over time can be a great insight into future outcomes.

```{r, echo = FALSE, message = FALSE, include = FALSE}
# load geographic data
get_congress_map <- function(cong=114) {
  tmp_file <- tempfile()
  tmp_dir  <- tempdir()
  zp <- sprintf("http://cdmaps.polisci.ucla.edu/shp/districts%03i.zip",cong)
  download.file(zp, tmp_file)
  unzip(zipfile = tmp_file, exdir = tmp_dir)
  fpath <- paste(tmp_dir, sprintf("districtShapes/districts%03i.shp",cong), sep = "/")
  st_read(fpath)
}

# load 114th congress data
cd114 <- get_congress_map(114)

# load house district data
districtvote <- read_csv("house party vote share by district 1948-2020.csv")

# filter districtvote set for selected variables
districtvote_2014 <- districtvote %>%
  filter(raceYear == 2014) %>%
  select(raceYear, State, DemVotes, RepVotes, district_num, district_id, WinnerParty, state_abb) %>%
  group_by(district_num) %>%
  mutate(RepVoteMargin = (RepVotes)/(RepVotes + DemVotes),
         DemVoteMargin = (DemVotes)/(RepVotes + DemVotes),
         VoteMargin = (RepVoteMargin - DemVoteMargin)) %>%
  rename(DISTRICT = district_num) %>%
  rename(STATENAME = State)

# checking which variable class to merge on
class(districtvote_2014$DISTRICT)

# checking which variable class to merge on
class(cd114$DISTRICT)

# need to match up these classes of variables so I can merge
# change cd114 variable class to numeric
cd114$DISTRICT <- as.numeric(cd114$DISTRICT)

# verifying new variable class
class(cd114$DISTRICT)

# join datasets based on DISTRICT
districtmargin <- cd114 %>%
  left_join(districtvote_2014,
            by = c("DISTRICT","STATENAME"))
head(districtmargin)

# To plot faster
districts_plot <- rmapshaper::ms_simplify(districtmargin, keep = 0.01)
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
# plot map
ggplot() +
  geom_sf(data = districts_plot,
          aes(fill = VoteMargin),
          inherit.aes = FALSE,alpha = 0.9) +
  scale_fill_gradient2(low = "Blue",
                       mid = "White",
                       high = "Red",
                       name = "Vote Share Margin") +
  coord_sf(xlim = c(-125.27, -66.57),
           ylim = c(20.55, 50.23),
           expand = FALSE) +
  labs(title = "Vote Margin by U.S. Congressional District, 2014",
       caption = "Data Provided by Professor Ryan Enos of Harvard\nUniversity along with Jeffrey B. Lewis of UCLA, et al.",
       subtitle = "A positive margin indicates a GOP victory.\nA negative margin indicates a Democratic Party victory.\nA margin larger in magnitude indicates a more decisive victory.\n") +
  theme_void() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())

# leaving out alaska and hawaii to better see variation in CD in 
# the contiguous US
# Did not include CD numbers on map because of hwo messy it gets
```

### Congressional Seat Share

Looking at Congressional Seat Share gets to where the real power lies in Congress: the more seats your party possesses, the more powerful you can be in your duties. For instance, the Democrats can win a district by 99 points or 1 point, and in either situation, they would send only one representative to the House. The ability of a party to win a district is a crucial element to their agenda. Getting 48 percent of the vote, which slims down the vote share margin of the winning party, nonetheless results in a lost election. For example, if the Democratic Party consistently gets 45% of the vote in a particular district's election, and thus losing each time, then that can serve as a good predictor that the GOP most likely will win future elections. Here we see that most states are heavily dominated by one party, with few residing right on the line of equal numbers of representatives from the GOP and Democratic Party. 

```{r, echo = FALSE, message = FALSE}
# load in data
popvote2 <- read_csv("house party vote share by district 1948-2020.csv")

# filter for national seat share in 2014
seatshare1 <- popvote2 %>%
  filter(raceYear == 2014) %>%
  select(State, raceYear, CD, WinnerParty) %>%
  # make a numeric variable showing a seat gain
  mutate(dem_win = ifelse(WinnerParty == "D", 1, 0),
         rep_win = ifelse(WinnerParty == "R", 1, 0)) %>% 
  group_by(State, raceYear) %>%
  # calculate individual seat shares and the seat share margin
  summarise(Dem_Seat_Share = (sum(dem_win)/sum(rep_win + dem_win)),
         Rep_Seat_Share = (sum(rep_win)/sum(rep_win + dem_win)),
         Seat_Margin = sum(Rep_Seat_Share - Dem_Seat_Share)) %>%
  rename("state" = State) 

# plot US Map Seat Share
seat_map <- plot_usmap(data = seatshare1,
                        regions = "states",
                        values = "Seat_Margin",
                        labels = TRUE,
                        label_color = "black") +
  scale_fill_gradient2(low = "blue",
                      mid = "white",
                      high = "red",
                      name = 'Seat Share Margin') +
  labs(title = "U.S. Congressional Seat Share by State, 2014",
       caption = "Data Provided by Professor Ryan Enos of Harvard University.",
       subtitle = "Positive values indicate a majority of House seats held by the Republican Party.\nNegative values indicate a majority of House seats held by the Democratic Party.\nA margin larger in magnitude indicates more seats held by the majority party.\n") +
  theme_void()

# Set labels font size
seat_map$layers[[2]]$aes_params$size <- 2.5
print(seat_map)
```

## Relating Back to Nebraska's 2nd Congressional District

### Nebraska's Living Bipartisanship

The culmination of my work over the next two months will result in a prediction of the election in Nebraska's 2nd Congressional District (NE-02). Thus, I would like to discuss Nebraska particular situation regarding vote share margins and seat shares. In the visualizations below, I have included data from 2014 (to compare to those above), 2018 (the most recent midterm election), and 2020 (the most recent voting data).

As of recently, Nebraska has garnered a public reputation of being a dependable state for the GOP. In the visualizations below, it is not hard to understand why that is --- especially when Nebraska has had 2 Republican senators throughout this time period as well. However, just under a decade ago, Nebraska had a Democratic senator, and as the visualization displays, a Democratic representative from NE-02. In this district lies Nebraska's foremost urban city, Omaha, and its sprawling suburbs, an environment becoming increasingly favorable to Democrats nationwide. While NE-02's Democratic elected official only lasted one term in office after his election in 2014, this phenomenon proved once more that there is potential for a Democrat to emerge from Nebraska.   

```{r, echo = FALSE, message = FALSE, include = FALSE}
# load in data for 2014 and 2020 Nebraska seat Margins

# load 114th congress data
cd114_ne2 <- get_congress_map(114)

# load house district data
districtseat_ne2 <- read_csv("house party vote share by district 1948-2020.csv")

# Facet Wrap: 2014 thru 2020 Nebraska GOP Seat Share
# filter districtvote set for selected variables
districtseat_ne2014thru2020 <- districtseat_ne2 %>%
  filter(raceYear %in% c(2014,2018,2020),
         State == "Nebraska") %>%
  select(State, raceYear, WinnerParty, district_num) %>%
  # make a numeric variable showing a seat gain
  mutate(WinnerParty = case_when(
    WinnerParty == "D" ~ "Democrat",
    WinnerParty == "R" ~ "Republican")) %>% 
  group_by(State, raceYear) %>%
  summarise(district_num,
            WinnerParty) %>%
  rename(DISTRICT = district_num) %>%
  rename(STATENAME = State)


# checking which variable class to merge on
class(districtseat_ne2014thru2020$DISTRICT)

# checking which variable class to merge on
class(cd114_ne2$DISTRICT)

# need to match up these classes of variables so I can merge
# change cd114 variable class to numeric
cd114_ne2$DISTRICT <- as.numeric(cd114_ne2$DISTRICT)

# verifying new variable class
class(cd114_ne2$DISTRICT)

# join datasets based on DISTRICT
districtseat_ne2014thru2020 <- cd114_ne2 %>%
  left_join(districtseat_ne2014thru2020,
            by = c("DISTRICT","STATENAME")) %>%
  filter(STATENAME == "Nebraska") %>%
  select(STATENAME, DISTRICT, raceYear, WinnerParty)
head(districtseat_ne2014thru2020)

# To plot faster
districtseat_plot_ne2014thru2020 <- rmapshaper::ms_simplify(districtseat_ne2014thru2020, keep = 0.01)
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot() +
  geom_sf(data = districtseat_plot_ne2014thru2020,
          aes(fill = WinnerParty),
          inherit.aes = FALSE,
          alpha = 0.9) +
  geom_sf_text(data = districtseat_plot_ne2014thru2020,
               aes(label = DISTRICT), 
               colour = "white",
               size = 2.25,
               check_overlap = TRUE) +
  facet_wrap(facets = raceYear ~.) +
  scale_fill_manual(values = c("blue", "red")) +
  labs(title = "Party Affiliations of Nebraska's House of Representatives Delegation\n",
       caption = "Data Provided by Professor Ryan Enos of Harvard\nUniversity along with Jeffrey B. Lewis of UCLA, et al.",
       fill = "Party Affiliation of Elected Representative") +
  theme_void() +
  theme(legend.position = "bottom",
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())
```

### Nebraska's Urban-Rural Divide

A common sight across the United States is how rural areas are sprinting toward the GOP while the urban and suburban spheres are huddling around the Democratic Party. In Nebraska, that seems to ring true, according to their vote share margin trends. In NE-01 lies Nebraska's capital city, Lincoln, with around [300,000 people.](https://www.census.gov/quickfacts/lincolncitynebraska) The vast majority of Nebraska's rural land constitutes NE-03, which presents an interesting case of gerrymandering, where this district's trend of an increasing GOP vote share margin almost guarantees a GOP win each cycle. 

```{r, echo = FALSE, message = FALSE, include = FALSE}
# load in data for 2014 through 2020 Nebraska Vote Margins

# load 114th congress data
cd114_ne3 <- get_congress_map(114)

# load house district data
districtvote_ne3 <- read_csv("house party vote share by district 1948-2020.csv")

# Facet Wrap: 2014 and 2020 Nebraska GOP Vote Share Margin
# filter districtvote set for selected variables
districtvote_ne2014thru2020 <- districtvote_ne3 %>%
  filter(raceYear %in% c(2014,2018, 2020),
         State == "Nebraska") %>%
  select(raceYear, State, DemVotes, RepVotes, district_num, district_id, WinnerParty, state_abb) %>%
  group_by(district_num) %>%
  # find vote share margins and total vote share margin
  mutate(RepVoteMargin = ((RepVotes)/(RepVotes + DemVotes)),
         DemVoteMargin = ((DemVotes)/(RepVotes + DemVotes)),
         VoteMargin = (RepVoteMargin - DemVoteMargin)) %>%
  rename(DISTRICT = district_num) %>%
  rename(STATENAME = State)


# checking which variable class to merge on
class(districtvote_ne2014thru2020$DISTRICT)

# checking which variable class to merge on
class(cd114_ne3$DISTRICT)

# need to match up these classes of variables so I can merge
# change cd114 variable class to numeric
cd114_ne3$DISTRICT <- as.numeric(cd114_ne3$DISTRICT)

# verifying new variable class
class(cd114_ne3$DISTRICT)

# join datasets based on DISTRICT
districtvote_ne2014thru2020 <- cd114_ne3 %>%
  left_join(districtvote_ne2014thru2020,
            by = c("DISTRICT","STATENAME")) %>%
  filter(STATENAME == "Nebraska") %>%
  select(STATENAME, DISTRICT, raceYear, DemVotes, RepVotes, DemVoteMargin, RepVoteMargin, VoteMargin)
head(districtvote_ne2014thru2020)

# To plot faster
districtvote_plot_ne2014thru2020 <- rmapshaper::ms_simplify(districtvote_ne2014thru2020, keep = 0.01)
```

```{r, echo = FALSE, message = FALSE, warning = FALSE}
ggplot() +
  geom_sf(data = districtvote_plot_ne2014thru2020,
          aes(fill = VoteMargin),
          inherit.aes = FALSE,alpha = 0.9) +
  facet_wrap(facets = raceYear ~.) +
  geom_sf_text(data = districtseat_plot_ne2014thru2020,
               aes(label = DISTRICT), 
               colour = "white",
               size = 2.25,
               check_overlap = TRUE) +
  scale_fill_gradient(low = "Purple",
                      high = "Red",
                      name = "Vote Share Margin") +
  labs(title = "GOP Vote Margin in Nebraska",
       caption = "Data Provided by Professor Ryan Enos of Harvard\nUniversity along with Jeffrey B. Lewis of UCLA, et al.",
       subtitle = "A positive margin indicates a GOP victory.\nA negative margin indicates a Democratic Party victory.\nA margin larger in magnitude indicates a more decisive victory.\n") +
  theme_void() +
  theme(legend.position = "bottom",
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())

#lost in NE01 and NE02 ish but gain in NE03
```

A look into NE-01 shows a suprisingly rapid realignment in favor of the Democratic Party: almost cutting in half their 40 point margin deficit in 4 years. This still leaves the district with a 2 election streak of a GOP victory by more than 20 points, signifying that a Republican still ought to be heavily favored solely on party affiliation.

NE-02 is where promise shows for Nebraskan Democrats. In addition to winning in 2014, the Democrats were able to sustain their close competition in the district over the next 6 years, eventually delivering one electoral vote to now-President Joe Biden. By looking into NE-02's history with their representative's party affiliation and vote share margins, it is not at all evident that one party is overly favored to win this district every two years. While that is not a prediction in favor of any one party or candidate, it does help to rule out the phenomenon of "one party rule" that has infected many districts across the country. Through this data analysis, I am led to believe that this district contains a competitiveness that gives a Democrat in Nebraska a real avenue toward a groundbreaking win. In the currnet political environment that revolves around an unpopular incumbent Democratic president in his fist term, a struggling economy, and abortion rights, this district will be a fantastic way to learn about how voters behave.

#### Notes:
[My project repository can be viewed here.](https://github.com/luke-tomes/election-forecast)



#### Sources:

Jeffrey B. Lewis, Brandon DeVine, Lincoln Pitcher, and Kenneth C. Martis. (2013) Digital Boundary Definitions of United States Congressional Districts, 1789-2012. Retrieved from https://cdmaps.polisci.ucla.edu on [09-15-2022].

House general elections, All States, 2014 summary. (2022). CQ voting and elections collection (web site). http://library.cqpress.com.ezp-prod1.hul.harvard.edu/elections/avg2014-3us1