---
title: Does Media Advertising Matter?
author: Luke Tomes
date: '2022-10-11'
slug: media-advertising
categories: []
tags: []
authors: []
description: ''
externalLink: ''
series: []
---

In this blog, I will seek to highlight how advertising in the airwaves can be used in predicting elections. To start, I will explore the relationships between the number of ads run in a district by each candidate and then use the 2018 advertising data as a stand-in, because I currently do not have the data from the current 2022 cycle, for my 2022 midterm elections prediction.

```{r setup, echo = FALSE, message = FALSE, include = FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE,
                      message = FALSE,
                      warning = FALSE,
                      fig.align = "center")
```

```{r libraries, include = FALSE}
# Load Libraries
library(tidyverse)
library(usmap)
library(sf)
library(rmapshaper)
library(car)
library(moderndive)
library(janitor)
```

```{r load data, include = FALSE}
# load existing datasets for model
polls_rdi_election_vote_filter2008 <- read_csv("polls_rdi_election_vote_filter2008.csv")
polls_rdi_election_vote_filter2008_sum <- read_csv("polls_rdi_election_vote_filter2008_sum.csv")
polls_rdi_election_seat_filter2008 <- read_csv("polls_rdi_election_seat_filter2008.csv")
polls_rdi_election_seat_filter2008_sum <- read_csv("polls_rdi_election_seat_filter2008_sum.csv")
district_by_district_model_data <- read_csv("district_by_district_model_data.csv")

# load datasets for extension
ads_compress <- readRDS("ads_2006_2018.rds")
ads_issues_compress <-readRDS("ads_issues_2012-2018.rds")
state_and_districts <- read_csv("State and district names.csv")

# data from last week
house_party_vote_share_by_district_1948_2020 <- read_csv("house party vote share by district 1948-2020.csv")
all_expert_ratings_2018 <- read_csv("2018_ratings_share.csv")
attached_data <- read_csv("attached_data.csv")
```

```{r prep datasets from last week}
# edit election dataset 
house_party_vote_share_by_district_1948_2020 <- house_party_vote_share_by_district_1948_2020 %>%
  mutate(RepStatus = case_when(
    DemVotesMajorPercent == 100.00 ~ "Challenger",
    TRUE ~ RepStatus),
    DemStatus = case_when(
    RepVotesMajorPercent == 100.00 ~ "Challenger",
    TRUE ~ DemStatus),
    RepStatus = case_when(
      RepCandidate == "Rothfus, Keith" & raceYear == 2018 ~ "Incumbent",
      TRUE ~ RepStatus),
    DemStatus = case_when(
      DemCandidate == "Lamb, Conor" & raceYear == 2018 ~ "Challenger",
      TRUE ~ DemStatus))
    
    
# wrangle vote dataset for 2016 seat party affiliations
seat_party_2016 <- house_party_vote_share_by_district_1948_2020 %>%
  filter(raceYear == 2016) %>%
  mutate(party_2016 = WinnerParty) %>%
  select(party_2016, State, district_num)

# wrangle vote dataset for 2018
vote_data_wrangle <- house_party_vote_share_by_district_1948_2020 %>%
  filter(raceYear == 2018) %>%
  left_join(seat_party_2016, by = c("State" = "State", 
                                    "district_num" = "district_num")) %>%
  mutate(district_num = as.character(district_num)) %>%
  select(State, raceYear, PluralityParty, RepVotesMajorPercent, DemVotesMajorPercent, state_abb, RepStatus, DemStatus, CD, district_num, district_id, WinnerParty, vote_margin, R_vote_margin, party_2016) %>%
  rename(year = raceYear,
         district = district_num)

# Prep all-district 2018 expert ratings
edit_all_expert_ratings_2018 <- all_expert_ratings_2018 %>%
  mutate(State = as.character(State),
         district_num = as.character(district_num)) %>%
  mutate(district_num = case_when(
           district_num == "AL" ~ "0",
           TRUE ~ district_num)) %>%
  select(-WinnerParty, -District, -CD, -state_abb, -district_id, -year) %>%
  rename(State = State,
         district = district_num)

# combine datasets
ratings_polls_2018 <- vote_data_wrangle %>% 
  left_join(edit_all_expert_ratings_2018, by = c("State" = "State", "district" = "district")) %>%
  mutate(RepStatus = case_when(
    DemVotesMajorPercent == 100.00 ~ "Challenger",
    TRUE ~ RepStatus),
    DemStatus = case_when(
    RepVotesMajorPercent == 100.00 ~ "Challenger",
    TRUE ~ DemStatus),
    D_vote_margin = DemVotesMajorPercent - RepVotesMajorPercent,
    winner_status = case_when(
      WinnerParty == "R" ~ RepStatus,
      WinnerParty == "D" ~ DemStatus),
    open_seat = case_when(
      DemStatus == "Challenger" & RepStatus == "Challenger" ~ "Yes",
      TRUE ~ "No"),
    open_seat = case_when(
      district_id == "MA07" ~ "No",
      district_id == "NY14" ~ "No",
      district_id == "NC09" ~ "No",
      district_id == "SC01" ~ "No",
      TRUE ~ open_seat),
    incumbent_winner = case_when(
           # 1 = incumbent won
           # 0 = incumbent lost
      winner_status == "Incumbent" ~ 1,
      winner_status == "Challenger" ~ 0),
    # take out ME-02 duplicate
    take_out = case_when(
      vote_margin == 0.008155614 ~ "yes",
      TRUE ~ "no"),
    incumbent_winner = as.character(incumbent_winner),
    Hold_or_Gain = case_when(
      WinnerParty == "R" & party_2016 == "R" ~ "Republican Hold",
      WinnerParty == "R" & party_2016 == "D" ~ "Republican Gain",
      WinnerParty == "D" & party_2016 == "D" ~ "Democratic Hold",
      WinnerParty == "D" & party_2016 == "R" ~ "Democratic Gain",)) %>%
  filter(take_out != "yes") %>%
  select(-take_out) %>%
  distinct(CD, .keep_all= TRUE) %>%
  rename(DISTRICT = district) %>%
  rename(STATENAME = State)

# Show the seat share expert picks would result in
ratings_polls_2018_experts <- ratings_polls_2018 %>%
  mutate(expert_rating = case_when(
    avg <= 2 ~ "Noncompetitive",
    avg <= 2.75  ~ "Likely Dem",
    avg <= 3.5 ~ "Lean Dem",
    avg <= 4 ~ "Dem Toss-up",
    avg == 4 ~ "Toss-up",
    avg <= 4.5 ~ "GOP Toss-up",
    avg <= 5.25 ~ "Lean GOP",
    avg <= 6 ~ "Likely GOP",
    avg <= 7 ~ "Noncompetitive"))
```

```{r data prep for plot for expert predictions, warnings = FALSE, include = FALSE}
# PLOT EXPERT RATINGS
  
# load geographic data
get_congress_map <- function(cong=114) {
  tmp_file <- tempfile()
  tmp_dir  <- tempdir()
  zp <- sprintf("http://cdmaps.polisci.ucla.edu/shp/districts%03i.zip",cong)
  download.file(zp, tmp_file)
  unzip(zipfile = tmp_file, exdir = tmp_dir)
  fpath <- paste(tmp_dir, sprintf("districtShapes/districts%03i.shp",cong), sep = "/")
  st_read(fpath)
}

# load 114th congress data
cd114 <- get_congress_map(114)

# checking which variable class to merge on
class(ratings_polls_2018$DISTRICT)

# checking which variable class to merge on
class(cd114$DISTRICT)

# need to match up these classes of variables so I can merge
# change cd114 variable class to numeric
ratings_polls_2018$DISTRICT <- as.character(ratings_polls_2018$DISTRICT)

# verifying new variable class
class(ratings_polls_2018$DISTRICT)

# join datasets based on DISTRICT
district_map_experts <- cd114 %>%
  left_join(ratings_polls_2018_experts,
            by = c("DISTRICT",
                   "STATENAME"))

# To plot faster
district_plot_experts <- rmapshaper::ms_simplify(district_map_experts, keep = 0.01)
```

#### How Ad Money is Distirbuted

[Professor Ryan D. Enos](https://hollis.harvard.edu/primo-explore/fulldisplay?docid=TN_cdi_proquest_miscellaneous_59786011&context=PC&vid=HVD2&search_scope=everything&tab=everything&lang=en_US) has detailed that not all congressional districts run advertisements over the airways --- and that is because many districts throughout the country simply are noncompetitive. When one can predict the outcome of an election in a district with relative confidence, there is no need for an advertising blitz that costs an enormous amount of money. However, in the districts that are competitive, fundraising can run wild, allowing for expensive advertising over the airways. Below is a map of the competitive districts in the 2018 cycle (and notice how few of them there are), with the districts that have a rating of "Solid Dem/Solid GOP" renamed as "Noncompetitive."


```{r plot for expert predictions}
# plot 2018 seat share
ggplot() +
  geom_sf(data = district_plot_experts,
          aes(fill = expert_rating),
          inherit.aes = FALSE,
          alpha = 1) +
  scale_fill_manual(values = c("#e6e7ff", # Dem Toss up
                               "#fff1f0", # GOP Toss up
                               "#c4d8ff", # Lean Dem
                               "#ffcfcf", # Lean GOP
                               "#6ea1ff", # Likely Dem
                               "#ff4040", # Likely GOP
                               "Gray", # Noncompetitive
                               "White"), # Toss up,
                   name = "Predictions") +
  coord_sf(xlim = c(-125.27, -66.57),
           ylim = c(23.75, 50.23),
           expand = FALSE) +
  labs(title = "Expert Seat Predictions by U.S. Congressional District, 2018",
       caption = "Data Provided by Professor Ryan Enos of Harvard\nUniversity along with Jeffrey B. Lewis of UCLA, et al.") +
  guides(fill = guide_legend(ncol = 1)) +
  theme_void() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())
```

#### Concentration of Ads in the 2018 Cycle

Competition within a district, especially when a district is seem as a "toss-up" battleground between the Democrat and Republican, attracts donor money like no other. Thus, one would expect the districts with lighter shades of their respective color to have more advertisements ran in their cycle than those with darker shades (less competitive). Our graph below confirms the notion that more competition results in larger numbers of advertisements. This is evidenced by competitive districts in states such as Iowa and Minnesota and in regions like the Northeast. Also we can see just how few, if any, advertisements are run in noncompetitive districts.


```{r ad numbers by district data wrangling}
# Make an at-large district = 0
ads_compress <- ads_compress %>%
  mutate(district = case_when(
    district == "1" & state == "Alaska" ~ "0",
    district == "1" & state == "Delaware" ~ "0",
    district == "1" & state == "Montana" ~ "0",
    district == "1" & state == "North Dakota" ~ "0",
    district == "1" & state == "South Dakota" ~ "0",
    district == "1" & state == "Vermont" ~ "0",
    district == "1" & state == "Wyoming" ~ "0",
    TRUE ~ district
  ))


# Organize total ads for democrats in 2018
ads_total_2018 <- ads_compress %>%
  filter(cycle == "2018") %>%
  select(-creative, -dma, st_fips, cd_fips, st_cd_fips) %>%
  count(state, district) %>%
  rename(total_ads = n) %>%
  drop_na(state, district) %>%
  mutate(state = as.character(state),
         district = as.character(district)) %>%
  rename(STATENAME = state,
         DISTRICT = district) 

# Organize total ads for democrats in 2018
ads_dems_2018 <- ads_compress %>%
  filter(party == "Democrat",
         cycle == "2018") %>%
  select(-creative, -dma, st_fips, cd_fips, st_cd_fips) %>%
  count(state, district) %>%
  rename(dem_ads = n) %>%
  drop_na(state, district) %>%
  mutate(state = as.character(state),
         district = as.character(district)) %>%
  rename(STATENAME = state,
         DISTRICT = district)

# Organize total ads for democrats in 2018
ads_gop_2018 <- ads_compress %>%
  filter(party == "Republican",
         cycle == "2018") %>%
  select(-creative, -dma, st_fips, cd_fips, st_cd_fips) %>%
  count(state, district) %>%
  rename(gop_ads = n) %>%
  drop_na(state, district) %>%
  mutate(state = as.character(state),
         district = as.character(district)) %>%
  rename(STATENAME = state,
         DISTRICT = district)

# Find difference in number of ads in 2018
ads_diff_2018 <- ads_dems_2018 %>%
  left_join(ads_gop_2018, by = c("STATENAME", "DISTRICT")) %>%
  drop_na(dem_ads, gop_ads) %>%
  mutate(ads_diff = dem_ads - gop_ads) %>%
  mutate(STATENAME = as.character(STATENAME),
         DISTRICT = as.character(DISTRICT))
```

```{r plot total number of ads}
# Plot total number of ads

# join datasets based on DISTRICT
district_map_total_ads <- cd114 %>%
  left_join(ads_total_2018,
            by = c("DISTRICT",
                   "STATENAME")) %>%
  mutate(total_ads = as.numeric(total_ads))

# To plot faster
district_plot_total_ads <- rmapshaper::ms_simplify(district_map_total_ads, keep = 0.01)

# plot 2018 seat share
ggplot() +
  geom_sf(data = district_plot_total_ads,
          aes(fill = total_ads),
          inherit.aes = FALSE,
          alpha = 1) +
  scale_fill_gradient(high = "red",
                      low = "white",
                      limits = c(1,43000),
                      name = "Number of Ads") +
  coord_sf(xlim = c(-125.27, -66.57),
           ylim = c(23.75, 50.23),
           expand = FALSE) +
  labs(title = "Total Ads Run in Each U.S. Congressional District, 2018",
       caption = "Data Provided by Professor Ryan Enos of Harvard\nUniversity along with Jeffrey B. Lewis of UCLA, et al.") +
  theme_void() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())
```

#### Which Party Holds the Advantage in the Airwaves

Despite all these ads, one has to wonder whether they advantage one candidate or another. Thus, a way to investigate whether a party gains an advantage in the airwaves is by comparing who runs more ads. Using the total number of ads by party may not allow us to view the entire picture, because if a Democrat runs 2000 advertisements in one cycle, the effect of those ads could be totally negated by 2000 ads run by the Republican opponent [as discussed by Ryan D. Enos.](https://www.ryandenos.com)

Thus, I believe finding the difference between ads run by the Democratic and Republican candidates will allow us to see who has the advantage in the airwaves. With the map and plot of the 2018 cycle below, we see the Democrats dwarfed the Republicans in advertisements. The big question that remains, though, is whether this strategy to run many more advertisements pays off with a win.


```{r plot difference in number of ads}
# Plot total number of ads

# join datasets based on DISTRICT
district_map_diff_ads <- cd114 %>%
  left_join(ads_diff_2018,
            by = c("DISTRICT",
                   "STATENAME")) %>%
  mutate(ads_diff = as.numeric(ads_diff))

# To plot faster
district_plot_diff_ads <- rmapshaper::ms_simplify(district_map_diff_ads, keep = 0.01)

# plot 2018 seat share
ggplot() +
  geom_sf(data = district_plot_diff_ads,
          aes(fill = ads_diff),
          inherit.aes = FALSE,
          alpha = 1) +
  scale_fill_gradient2(high = "Light Blue",
                       mid = "white",
                       low = "Dark Red",
                       name = "Difference in Ads") +
  coord_sf(xlim = c(-125.27, -66.57),
           ylim = c(23.75, 50.23),
           expand = FALSE) +
  labs(title = "Difference in Ads Run by U.S. Congressional District, 2018",
       subtitle = "Blue = Dem Advantage\nRed = GOP Advantage",
       caption = "Data Provided by Professor Ryan Enos of Harvard\nUniversity along with Jeffrey B. Lewis of UCLA, et al.") +
  guides(fill = guide_legend(ncol = 1)) +
  theme_void() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank())
```

```{r find total ads run by Dems and GOP}
ads_diff_2018_final <- ads_diff_2018 %>%
  mutate(ads_diff = as.numeric(ads_diff),
    advantage = case_when(
    ads_diff > 0 ~ "Democratic Advantage",
    ads_diff < 0 ~ "Republican Advantage",
    ads_diff == 0 ~ "No Party Advantage"
  ))

ggplot(data = ads_diff_2018_final,
       mapping = aes(x = ads_diff,
                     fill = advantage)) +
  geom_histogram(color = "white",
                 bins = 45) +
  scale_fill_manual(values = c("Blue", " Red"),
                    name = "Party Advantage") +
  geom_vline(xintercept = 0, lty = 2) +
  theme(legend.position = "bottom") +
  labs(title = "Democrat Advertisement Advantage by District",
       subtitle = "Positive Values Indicate Districts Where Democrats Ran More Ads")
```

#### Does an Ad Advantage Lead to Victory? 

[Scholars Gregory A. Huber and Kevin Arceneaux](https://hollis.harvard.edu/primo-explore/fulldisplay?docid=TN_cdi_proquest_miscellaneous_59786011&context=PC&vid=HVD2&search_scope=everything&tab=everything&lang=en_US) detail how presidential advertisements do not result in increased mobilization or in increasingly educated voters. However, they did find evidence that voters are persuaded by advertisements. Thus, one might expect the Democrats to perform extremely well in the districts where they enjoy an ad advantage over the Republicans. However, below in the map of 2018 districts, we see that advertisement advantages enjoyed by one party do not seem to ensure victory, and that an ad advantage is roughly unpredictive of election results.


```{r prep code for seeing if advantage led to win in 2018, warnings = FALSE, include = FALSE}
ad_win_match <- ratings_polls_2018 %>%
  left_join(ads_diff_2018, by = c("STATENAME", "DISTRICT")) %>%
  mutate(match = case_when(
    ads_diff > 0 & WinnerParty == "D" ~ "Won",
    ads_diff > 0 & WinnerParty == "R" ~ "Lost",
    ads_diff < 0 & WinnerParty == "R" ~ "Won",
    ads_diff < 0 & WinnerParty == "D" ~ "Lost"
  ))

# PLOT SEAT SHARE IN 2018

# checking which variable class to merge on
class(ad_win_match$DISTRICT)

# checking which variable class to merge on
class(cd114$DISTRICT)

# need to match up these classes of variables so I can merge
# change cd114 variable class to numeric
ad_win_match$DISTRICT <- as.character(ad_win_match$DISTRICT)

# verifying new variable class
class(ad_win_match$DISTRICT)

# join datasets based on DISTRICT
district_map_ad_match <- cd114 %>%
  left_join(ad_win_match,
            by = c("DISTRICT",
                   "STATENAME"))

# To plot faster
district_plot_ad_match <- rmapshaper::ms_simplify(district_map_ad_match,
                                         keep = 0.01)
```

```{r plot if advantage led to win in 2018}
# PLOT SEAT SHARE IN 2018
# plot 2018 seat share
ggplot() +
  geom_sf(data = district_plot_ad_match,
          aes(fill = match),
          inherit.aes = FALSE,
          alpha = 1) +
  scale_fill_manual(values = c("Red", 
                               "Green"),
                   name = "Result for Party Running More Ads") +
  coord_sf(xlim = c(-125.27, -66.57),
           ylim = c(23.75, 50.23),
           expand = FALSE) +
  labs(title = "Seat Map by U.S. Congressional District, 2018",
       caption = "Data Provided by Professor Ryan Enos of Harvard\nUniversity along with Jeffrey B. Lewis of UCLA, et al.",
       subtitle = "Democrats won 235 seats to the GOP's 200.\n") +
  guides(fill = guide_legend(ncol = 1)) +
  theme_void() +
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks.y=element_blank()) + 
  theme(legend.position = "bottom") +
  guides(fill = guide_legend(ncol = 2))
```

#### Historical Ad Advantage Trends

Here we compare ad advantages by district since the 2010 cycle, and what we see is a weak positive relationship between a Democrat's ad advantage and their ensuing vote share.


```{r historical data on Dem advertisment advantage and end result}
# Organize election dataset
house_data_set <- house_party_vote_share_by_district_1948_2020 %>%
  select(State, raceYear, RepVotesMajorPercent, DemVotesMajorPercent, district_num, R_vote_margin) %>%
  mutate(district_num = as.character(district_num)) %>%
  rename(DISTRICT = district_num,
         STATENAME = State,
         cycle = raceYear)

# Organize total ads for democrats in 2018
ads_dems_total <- ads_compress %>%
  filter(party == "Democrat") %>%
  select(-creative, -dma, st_fips, cd_fips, st_cd_fips) %>%
  count(state, district, cycle) %>%
  rename(dem_ads = n) %>%
  drop_na(state, district) %>%
  mutate(state = as.character(state),
         district = as.character(district)) %>%
  rename(STATENAME = state,
         DISTRICT = district)

# Organize total ads for democrats in 2018
ads_gop_total <- ads_compress %>%
  filter(party == "Republican") %>%
  select(-creative, -dma, st_fips, cd_fips, st_cd_fips) %>%
  count(state, district, cycle) %>%
  rename(gop_ads = n) %>%
  drop_na(state, district) %>%
  mutate(state = as.character(state),
         district = as.character(district)) %>%
  rename(STATENAME = state,
         DISTRICT = district)

# Find difference in number of ads in 2018
ads_diff_total <- ads_dems_total %>%
  left_join(ads_gop_total, by = c("STATENAME", "DISTRICT", "cycle")) %>%
  drop_na(dem_ads, gop_ads) %>%
  mutate(ads_diff = dem_ads - gop_ads) %>%
  mutate(STATENAME = as.character(STATENAME),
         DISTRICT = as.character(DISTRICT)) %>%
  mutate(DISTRICT = case_when(
    DISTRICT == "01" ~ "1",
    DISTRICT == "02" ~ "2",
    DISTRICT == "03" ~ "3",
    DISTRICT == "04" ~ "4",
    DISTRICT == "05" ~ "5",
    DISTRICT == "06" ~ "6",
    DISTRICT == "07" ~ "7",
    DISTRICT == "08" ~ "8",
    DISTRICT == "09" ~ "9",
    TRUE ~ DISTRICT)) %>%
  left_join(house_data_set, by = c("STATENAME", "DISTRICT", "cycle")) %>%
  drop_na(ads_diff,
          DemVotesMajorPercent)
```

```{r plot of historical data on Dem advertisment advantage and end result}
ggplot(data = ads_diff_total,
       mapping = aes(x = ads_diff,
                     y = DemVotesMajorPercent)) + 
  geom_point(alpha = 0.5,
             size = 1.5) +
  geom_smooth(method = "lm") +
  geom_vline(xintercept = 0, lty = 2, color = "red") +
  geom_hline(yintercept = 50, lty = 2, color = "red") + 
  xlim(-5000,8000) +
  ylim(20,75) +
  labs(title = "Democratic Vote Share vs Their Ad Advantage",
       subtitle = "Positive Values Indicate Districts Where Democrats Ran More Ads",
       x = "Democratic Ad Advantage",
       y = "Democratic Vote Share")
```

## Updated Prediction Model

Despite this weak correlation between a Democrat's ad advantage and their resulting vote share, I am going to incorporate it into my model alongside the national real disposable income percent change and national Democratic congressional ballot margin. This is because I want to see if one more predictor allows me to increase my predictive accuracy district-by-district.


```{r prep data for model}
# model dataset
model_data_1011 <- attached_data %>%
  left_join(ads_diff_total, by = c("state" = "STATENAME",
                                   "district" = "DISTRICT",
                                   "year" = "cycle")) %>%
  select(-RepVotesMajorPercent, - DemVotesMajorPercent, -R_vote_margin) %>%
  drop_na(ads_diff, dem_ads, gop_ads) %>%
  group_by(state, district) %>%
  filter(n() > 1) %>% # Filtering out single data rows
  group_nest() %>% 
  mutate(data = map(data, ~unnest(., cols = c())))

write_csv(model_data_1011, "model_data_10-11.csv")

# use 2018 numbers for 2022
data_model_2022 <- ads_diff_2018 %>% 
  rename(state = STATENAME,
         district = DISTRICT) %>%
  select(-dem_ads, -gop_ads) %>%
  mutate(mean_dem_margin = -0.9,
         rdi_change_pct = 1.91,
         year = 2022) %>%
  group_by(state, district) %>% 
  group_nest() %>% 
  mutate(data = map(data, ~unnest(., cols = c())))
```

```{r build out model}
# Building models
vote_model_1011 <- model_data_1011 %>% 
  mutate(model = map(data, ~lm(dem_votes_major_percent ~
                                 rdi_change_pct +
                                 mean_dem_margin +
                                 ads_diff,
                               data = .x))) %>% 
  select(-data)

# Extracting model results
vote_model_results_1011 <- vote_model_1011 %>% 
  mutate(r_squared = map_dbl(model, ~summary(.x)$r.squared))

# Predicting 2022 with a model
pred_vote_2022 <- data_model_2022 %>%
  # inner join as there may not be historical models for some districts
  inner_join(vote_model_1011, by = c("state", "district"))%>% 
  mutate(pred = map_dbl(.x = model,
                        .y = data, 
                        ~predict(object = .x, 
                                 newdata = as.data.frame(.y)))) %>%
  select(state, district, pred)
```

#### Results by Congressional District
```{r district predictions table}
print.data.frame(pred_vote_2022)
```

Here we the predicted vote shares of 77 districts. Some predictions are plausible, as many lie around 40-60 percent, which is where election results tend to fall for candidates. However, I am skeptical of this model, as my district of interest for this blog, NE-02, shows the Democratic candidate winning with 52.75. Historically, this district  has been ultra-competitive, but a Democrat has not achieved about 50% there in quite some time (and is predicted to be unlikely this cycle by most [major outlets.](https://ballotpedia.org/Nebraska%27s_2nd_Congressional_District_election,_2022))

Additionally, there are some critical pitfalls to this model that shows some districts boasting predicted vote shares above 100% and below 0%, both of which are impossible in a vote.


#### R-Squared Comparisons

Just like last week's forecast, the R-squared numbers for my predictions by district are graphed below, and while a higher R-squared is always desired, a value of 1 often should raise eyebrows in suspicion. Due to the sheer number that boast an R-squared value of 1, I am very pessimistic of this model and its effectiveness.


```{r vote share graph rsquared}
ggplot(data = vote_model_results_1011,
       mapping = aes(x = r_squared)) +
  geom_histogram()
```

#### Limitations of My Model

Therefore, after visualizing the trends and updating my model, I am not certain that the Democratic Ad Advantage variable should be included in my overall forecast. This blog resulted in me getting  smaller R-squared values and predicting less districts (due to less data being available on this front). 

A big note on advertisements that could explain why the aggregate variable may not be the best predictor is because the persuasive effects of advertisements have been shown to dissipate rapidly by scholar [Alan S. Gerber, et. al.](https://hollis.harvard.edu/primo-explore/fulldisplay?docid=TN_cdi_proquest_miscellaneous_881466543&context=PC&vid=HVD2&search_scope=everything&tab=everything&lang=en_US) Dr. Gerber goes on state that within a week or two after an advertisement has been run, its effects will have "all but disappeared." Therefore, by finding the difference between the total number of ads run by candidates in a cycle, we are accounting for ads run in May of an eleciton year. And frankly, those ads seem to wear off even before May turns into June. Thus, a better way to use a variable incorporating the number of ads would be to include only advertisements run two weeks before an election date.



##### Notes:

*This blog is part of a series of articles meant to progressively understand how election data can be used to predict future outcomes. I will add to this site on a weekly basis under the direction of [Professor Ryan D. Enos](https://www.ryandenos.com). In the run-up to the midterm elections on November 8, 2022, I will draw on all I have learned regarding what best forecasts election results, and I will predict the outcome of the NE-02 U.S. Congressional Election.*

[My project repository can be viewed here.](https://github.com/luke-tomes/election-blog)

##### Sources:

[Bureau of Economic Analysis, RDI Change](https://www.bea.gov/news/2022/personal-income-and-outlays-july-2022)
